---
title: "Darwin Core mapping"
subtitle: "For: Checklist of alien birds of Belgium"
author:
- Cristina Preda
- Tim Adriaens
- Peter Desmet
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/alien-birds-checklist/blob/master/src/dwc_mapping.Rmd).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
library(rgbif)          # To use GBIF services
```

# Read source data

The data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1ugsmAq_tlUjFmZef2zyw-mT0Tq1jSdPlNO4KTfJyXVo/edit#gid=0).

Read the relevant worksheet (published as csv):

```{r}
input_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSgGi_Un0-7cyg-SzaiE0-RYY5-WvuZNF9kG2GLgeonX6heR6U3xpechdMKWVMQ9raT6AuR86U_gQt9/pub?gid=0&single=true&output=csv", col_types = cols(.default = col_character()))
```

Copy the source data to the repository to keep track of changes:

```{r}
write_csv(input_data, here("data", "raw", "alien_bird_checklist_dump.csv"), na = "")
```

Preview data: 

```{r}
input_data %>% head()
```

# Preprocessing

## Tidy data

Clean names data somewhat:

```{r}
input_data %<>%
  remove_empty("rows") %>%    # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Scientific names

Use the [GBIF nameparser](https://www.gbif.org/tools/name-parser) to retrieve nomenclatural information for the scientific names in the checklist:

```{r}
parsed_names <- input_data %>%
  distinct(scientific_name) %>%
  pull() %>% # Create vector from dataframe
  parsenames() # An rgbif function
```

Show scientific names with nomenclatural issues, i.e. not of `type = SCIENTIFIC` or that could not be fully parsed (should be 0). Note: these are not necessarily incorrect.

```{r}
parsed_names %>%
  select(scientificname, type, parsed, parsedpartially, rankmarker) %>%
  filter(!(type == "SCIENTIFIC" & parsed == "TRUE" & parsedpartially == "FALSE"))
```

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom (that will remain the same as long as scientific name and kingdom remain the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "alien-birds-checklist", # e.g. "alien-fishes-checklist"
  "taxon",
  vdigest(scientific_name, algo = "md5"),
  sep = ":"
))
```

## Preview data

Preview data:

```{r}
input_data %>% head()
```
# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
taxon <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "INBO")
```

### accessRights

```{r}
taxon %<>% mutate(dwc_accessRights = "https://www.inbo.be/en/norms-data-use") 
```

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "https://doi.org/10.15468/wr3gis")
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "INBO")
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "Checklist of alien birds of Belgium") 
```

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name)
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = "Animalia")
```

### phylum

```{r}
taxon %<>% mutate(dwc_phylum = "Chordata")
```

### class

```{r}
taxon %<>% mutate(dwc_class = "Aves")
```

### order

```{r}
taxon %<>% mutate(dwc_order = order)
```

### family

```{r}
taxon %<>% mutate(dwc_family = family)
```

### genus

```{r}
taxon %<>% mutate(dwc_genus = genus)
```

### taxonRank

Inspect values:

```{r}
taxon %>%
  group_by(taxon_rank) %>%
  count()
```

Values are conform [GBIF rank vocabulary](http://rs.gbif.org/vocabulary/gbif/rank_2015-04-24.xml), map as is:

```{r}
taxon %<>% mutate(dwc_taxonRank = taxon_rank)
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = "ICZN")
```

# Distribution extension

Create a dataframe with all data:

```{r}
distribution <- input_data
```

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(dwc_taxonID = taxon_id)
```

### locationID

Inspect values:

```{r}
distribution %>%
  group_by(country_code, location) %>%
  count()
```

Set current `NA`'s in `location` to `Belgium`:

```{r}
distribution %<>% mutate(location = case_when(
  is.na(location) ~ "Belgium",
  TRUE ~ location
))
```

Map `locationID` to ISO 3166 code:

```{r}
distribution %<>% mutate(dwc_locationID = case_when(
  location == "Belgium" ~ "ISO_3166-2:BE",
  location == "Flanders" ~ "ISO_3166-2:BE-VLG",
  location == "Wallonia" ~ "ISO_3166-2:BE-WAL",
  location == "Brussels" ~ "ISO_3166-2:BE-BRU")
)
```

Inspect values:

```{r}
distribution %>%
  group_by(location, dwc_locationID) %>%
  count()
```

### locality

Map to standard values:

```{r}
distribution %<>% mutate(dwc_locality = recode(location, 
  "Flanders" = "Flemish Region",
  "Wallonia" = "Walloon Region",
  "Brussels" = "Brussels-Capital Region"
))
```

Inspect values:

```{r}
distribution %>%
  group_by(location, dwc_locality) %>%
  count()
```

### countryCode

Inspect values:

```{r}
distribution %>%
  group_by(country_code) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_countryCode = country_code) 
```

### occurrenceStatus

Inspect values:

```{r}
distribution %>%
  group_by(occurrence_status) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceStatus = occurrence_status) 
```

### establishmentMeans

```{r}
distribution %<>% mutate(dwc_establishmentMeans = "introduced")
```

### degreeOfEstablishment

Inspect values:

```{r}
distribution %>% 
  group_by(degree_of_establishment) %>% 
  count()
```

These categories of `degree_of_establishment` are conform the [Darwin Core Degree of Establishment vocabulary](http://rs.gbif.org/vocabulary/dwc/degree_of_establishment_2022-02-02.xml).

```{r}
distribution %<>% mutate(dwc_degreeOfEstablishment = degree_of_establishment) 
```

### pathway

Inspect values:

```{r}
distribution %>% 
  group_by(introduction_pathway) %>% 
  count()
```

```{r}
distribution %<>% mutate(dwc_pathway = introduction_pathway)
```

### eventDate

Inspect values for `date_first_observation`:

```{r}
distribution %>%
  group_by(date_first_observation) %>%
  count()
```

All date information should comply to the ISO 8601 standard, which requires a four-digit year as a minimum.
We here transform deviating values:

```{r}
distribution %<>% mutate(
  date_first_observation = recode(
    .x = date_first_observation,
    "18th century" = "1701",
    "1890s" = "1890",
    "1950ies" = "1950"
))
```

Inspect values for `date_last_observation`:

```{r}
distribution %>%
  group_by(date_last_observation) %>%
  count()
```

Inspect all combinations of `date_first_observation` and `date_last_observation`:

```{r}
distribution %>%
  group_by(date_first_observation, date_last_observation) %>%
  count()
```

Map `eventDate`:

```{r}
distribution %<>% mutate(
  dwc_eventDate = case_when(
    is.na(date_first_observation) & is.na(date_last_observation) ~ "",
    is.na(date_first_observation) & !is.na(date_last_observation) ~ paste(
      "", 
      date_last_observation, 
      sep = "/"),
    !is.na(date_first_observation) & is.na(date_last_observation) ~ paste(
      date_first_observation, 
      "", 
      sep = "/"),
    !is.na(date_first_observation) & !is.na(date_last_observation) ~ paste(
      date_first_observation, 
      date_last_observation, 
      sep = "/"
    )
  )
)
```

Show mapping:

```{r}
distribution %>%
  group_by(date_first_observation, date_last_observation, dwc_eventDate) %>%
  count()
```

### source

Inspect values:

```{r}
distribution %>%
  group_by(source) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_source = source) 
```

### occurrenceRemarks

Inspect values:

```{r}
distribution %>%
  group_by(remarks) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceRemarks = remarks) 
```

# Species profile extension

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
species_profile <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

### taxonID

```{r}
species_profile %<>% mutate(dwc_taxonID = taxon_id)
```

Inspect `realm`:

```{r}
species_profile %>% 
  group_by(realm) %>% 
  count()
```

### isMarine

```{r}
species_profile %<>% mutate(dwc_isMarine = "FALSE")
```

### isTerrestrial

```{r}
species_profile %<>% mutate(dwc_isTerrestrial = case_when(
  realm == "terrestrial" ~ "TRUE",
  TRUE ~ "FALSE"
))
```

### isFreshwater

```{r}
species_profile %<>% mutate(dwc_isFreshwater = case_when(
  realm == "freshwater" ~ "TRUE",
  TRUE ~ "FALSE"
))
```

# Description extension

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

Create `description`:

```{r}
description <- input_data
```

In the description extension we want to include the native range of the species.
A single taxon can have multiple native ranges of the same type, expressed as multiple rows in the description extension. The mapping will be as following:

- `type` = `native range`
- `description` = the specific content of the field `native_range`
- `language` = `en`

## Term mapping

### taxonID

```{r}
description %<>% mutate(dwc_taxonID = taxon_id)
```

Inspect values:

```{r}
description %>% 
  group_by(native_range) %>% 
  count()
```

Separate content using the pipe symbol as a separator:

```{r}
description %<>% separate(
  native_range,
  into = c("native_range_1", "native_range_2", "native_range_3", "native_range_4", "native_range_5"),
  sep = "\\s\\|\\s",
  remove = FALSE
)
```

Change from a wide to a long dataset and trim:

```{r}
description %<>% gather(key, value, starts_with("native_range_"), na.rm = TRUE) %>%
  mutate(value = str_trim(value))
```

Inspect unique values:

```{r}
description %>% 
  group_by(value) %>% 
  count()
```

Map to the [WGSRPD vocabulary](http://grassworld.myspecies.info/sites/grassworld.myspecies.info/files/tdwg_geo2.pdf):

```{r}
description %<>% mutate(dwc_description = recode(value,
  "Africa" = "Africa (WGSRPD:2)",
  "Australia" = "Australia (WGSRPD:50)",
  "Caribbean" = "Caribbean (WGSRPD:81)",
  "Central America" = "Central America (WGSRPD:80)",
  "China" = "China (WGSRPD:36)",
  "East Asia" = "Eastern Asia (WGSRPD:38)",
  "Eastern Asia" = "Eastern Asia (WGSRPD:38)",
  "Eastern Europe" = "Eastern Europe (WGSRPD:14)",
  "Europe" = "Europe (WGSRPD:1)",
  "Galapagos" = "Galapagos (WGSRPD:GAL)",
  "Hawaiian Islands" = "Hawaiian Islands (WGSRPD:HI)",
  "India" = "Indian Subcontinent (WGSRPD:40)",
  "Indonesia" = "Indonesia (WGSRPD:ID)",
  "Japan" = "Japan (WGSRPD:JAP)",
  "Madagascar" = "Madagascar (WGSRPD:MDG)",
  "New Zealand" = "New Zealand (WGSRPD:51)",
  "North America" = "Northern America (WGSRPD:7)",
  "Northern America" = "Northern America (WGSRPD:7)",
  "Panama" = "Panama (WGSRPD:PAN)",
  "Philippines" = "Philippines (WGSRPD:PHI)",
  "Portugal" = "Portugal (WGSRPD:POR)",
  "South America" = "Southern America (WGSRPD:8)",
  "Southeast Europe" = "Southeastern Europe (WGSRPD:13)",
  "Southern Africa" = "Southern Africa (WGSRPD:27)",
  "Spain" = "Spain (WGSRPD:SPA)",
  "Tanzania" = "Tanzania (WGSRPD:TAN)",
  "Western Asia" = "Western Asia (WGSRPD:34)"
))
```

Inspect mapped values:

```{r}
description %>% 
  group_by(value, description) %>% 
  count()
```

Add `type` of description:

```{r}
native_range %<>% mutate(type = "native range")
```

Select `taxon_id`, `description` and `type`:

```{r}
native_range %<>% select(taxon_id, description, type)
```

## Pathway of introduction
 


Select `taxon_id`, `description` and `type`:

```{r}
pathway %<>% select(taxon_id, description, type)
```


Select `taxon_id`, `description` and `type`:

```{r}
degree_of_establishment %<>% select(taxon_id, description, type)
```

Union native range, pathway of introduction and degree of establishment into a single description extension:

```{r}
description <- bind_rows(native_range, pathway, degree_of_establishment)
```

### taxonID

```{r}
description %<>% mutate(dwc_taxonID = taxon_id)
```

### description

```{r}
description %<>% mutate(dwc_description = description)
```

### type

```{r}
description %<>% mutate(dwc_type = type)
```

### language

```{r}
description %<>% mutate(dwc_language = "en")
```

# Post-processsing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
distribution %<>% select(starts_with("dwc_"))
species_profile %<>% select(starts_with("dwc_"))
description %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
colnames(species_profile) <- str_remove(colnames(species_profile), "dwc_")
colnames(description) <- str_remove(colnames(description), "dwc_")
```

Sort on `taxonID` (to maintain some consistency between updates of the dataset):

```{r}
taxon %<>% arrange(taxonID)
distribution %<>% arrange(taxonID)
species_profile %<>% arrange(taxonID)
description %<>% arrange(taxonID)
```

Preview taxon core:

```{r}
taxon %>% head()
```

Preview distribution extension:

```{r}
distribution %>% head()
```

Preview species profile extension:

```{r}
species_profile %>% head()
```

Preview description extension:

```{r}
description %>% head(10)
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```
